---
title: "STAT 2450 Assignment 2"
author: "Alice Liu"
date: 'Banner:  B00783546'
output:
  html_document: default
  pdf_document: default
  word_document: default
---
In the following exercises, use "for" loops to do your iteration.  
(Do NOT use the built in function "sum").

1. Use a "for" loop to evaluate the following sum, when x=.5. 

$$y = \sum_{j=0}^{10} x^j$$

```{r}
x=.5
y=0 # initialize y to 0
for (i in 0:10) {
y = y + x^i

}
print(y)
```

Your answer should be 1.999023.

(*5 points*)

2.  Use a for loop, and the built in function "sqrt",  to do the following:

    loop over the elements of x=c(16,-9,9,4,-1,0)
    
    if the element x[i] is positive, print x[i] and the square root of x[i]
      or else if x[i] is negative, print("x[i] is negative")

```{r}
x=c(16,-9,9,4,-1,0)
for (i in x){
if (i<0) { print("x[i] is negtive")
  } else 
     {print (c(i,sqrt(i)))}
     }

```

Your output should be equivalent to:

 16  4
 
 "x[i] is negative"
 
 9 3
 
 4 2
 
 "x[i] is negative"
 
 0 0



(*5 points*)


3.  Use a "for" loop to iterate over the elements of a vector x, and increment a counter
nodd when the element of x is odd.  Then print the number of even and odd elements of x.

Recall that an integer y is odd if  y%%2==1, and otherwise y is even.

Use the R commands below to generate a vector "x" of length 100.

```{r}
set.seed(10) 
x=sample(1:10,100,replace=T)
x
```

Then modify the following code.
```{r}
nodd=0  #initialize the counter nodd, which counts the number of odd elements of x
for(i in x){#for ( i in x ) {increment nodd in the loop when the associated element of x is odd}
  if(i%%2==1){
   nodd=nodd+1
  }
}
neven=length(x)-nodd #neven=calculate neven using the length of x and nodd.
paste("number of odd elements = ",nodd)
paste("number of even elements= ",neven)#paste("number of even elements = ",neven)
```
For this vector x, the number of odd elements should be 56.


(*5 points*)

4.  generate a random 5x5 matrix whose entries are the numbers 1,2, ... 25, but in random positions, using the following code:
 
```{r}
set.seed(27) #set the seed for the random number generator
x=matrix(sample(1:25), byrow=T,ncol=5) #
x
```


As you can see, on my system, I get the following input matrix:

<pre>
     [,1] [,2] [,3] [,4] [,5]
     
[1,]    5   18   22    9    8

[2,]   19   16    1   24   23

[3,]   17   25    3   11   21

[4,]   15   10    6    7    4

[5,]   13   20   14   12    2
</pre>

It does not matter if you have a different matrix. Just use your matrix
for the next exercise.



Then, using a pair of nested for loops, loop over the positions in the
matrix x, and if the associated element of x is odd, replace the element
by its negative.

```{r}
#print(x)
for(i in 1:dim(x)[1]){#for loop involving i {
   for (j in 1:dim(x)[2]) {#  for loop involving j {
       x[i,j]=ifelse(x[i,j]%%2==1, -x[i,j], x[i,j])#    x[i,j]=ifelse(x[i,j]%%2==1, -x[i,j], something here for odd elements)
   }
}
print(x)#print(x)
```


Your answer should be equivalent to the following answer (which is valid for the input matrix shown above)


<pre>
   -5    18   22   -9    8
   
   -19   16   -1   24  -23
   
   -17  -25   -3  -11  -21
   
   -15   10    6   -7   -4
   
   -13   20   14   12    2

</pre>

Use your own input matrix instead, if different.

(*5 points*)
