---
title: "STAT 2450 Assignment 3 (40 points) "
author: "Alice Liu"
date: 'Banner:  B00783546'
output:
  pdf_document: default
  html_document: default
  word_document: default
---
1.  Write a function to calculate miles per gallon given kilometres travelled,
and litres of gasonline used.  The function should have two arguments, litres
and kilometres, and should return the mileage in mpg.

```{r}
mpg=function(litres,kilometres){#mpg=function(  arguments here){
   miles=kilometres*5/8#mpg=
   gallons=litres/4.55
   mpg=miles/gallons
   return(mpg) #return(mpg)
}
mpg(litres=8.3,kilometres=100)
 

```

Test your function using input values of 100 kilometres and 8.3 litres.


(5 points)


2.  The roots of the quadratic $a x^2 + b x + c$ are given by

$$\frac{-b \pm \sqrt{b^2-4ac}}{2a}$$

If $b^2-4ac <0$, the quadratic has no real roots.

Write a function to calculate the real roots of a quadratic.  The function
should have 3 arguments, *a*, *b* and *c*.
If $b^2-4ac <0$, the function should print "quadratic has no real roots",
and then return(NULL).
Otherwise, the function should return a vector of length 2, those being
the real roots (which may be the same if  $b^2-4ac =0$).

Test your function using the quadratic $x ^2 - 3 x + 2$.

```{r}
root=function(a,b,c){
   d=b^2-4*a*c
   if(d<0){
   print("quadratic has no real roots")
   return(NULL)
   }
   root=(-b+c(-1,1)*sqrt(d))/(2*a)
   return(root)
}
root(1,-3,2)
 

```
(5 points)

3.  Where $x_1, x_2, \ldots , x_n$ is a sample from a normal distribution
with unknown mean $\mu$ and unknown variance $\sigma^2$, the level
$100(1-\alpha)$% confidence interval for $\mu$ is given by

$$\bar x \pm t_{1-\alpha/2, n-1} \frac{s}{\sqrt n}$$

where $\bar x$ and $s$ are the sample mean and sample standard deviation
of the data, and $t_{1-\alpha/2, n-1}$ cuts off an area $1-\alpha/2$ to
its left under the $t$ curve with $n-1$ degrees of freedom.

Write a function which has two arguments, a vector of data $x$, and
alpha, which should have a default value of .05.
The function should return
a vector of length 2, which contains the endpoints of the confidence interval.

The percentiles of the t-distrubtion can be calculated as follows. Suppose
that you want the 97.5'th percentile of the t-distribution with 23 degrees
of freedom.  This can be calculated in R as

```{r}
qt(.975,23)
```

Test your function by calculating the 99% confidence interval using
the following data

```{r}
set.seed(87612345)
data=rnorm(25,mean=4.5,sd=.75)
```
You can check your calculation using

```{r}
t.test(data,conf.level=.99)
```

When putting your two endpoints together, you may find something similar to the 
following to be useful.

```{r}
1+c(-1,1)*.25
```
```{r}
t_interval=function(data,alpha=.05){
    n=length(data)
    t=qt(1-alpha/2,n-1)
    xbar=mean(data)
    s=sd(data)
    t_interval=xbar+c(-1,1)*t*s/sqrt(n)
return(t_interval)
}
t_interval(data,.01)
```
(5 points)

4.
The derivative of a function $f(x)$ can be approximated by the Newton's quotient

$$\frac{f(x+h) - f(x)}{h}$$

where $h$ is a small number.  Write a function to calculate the Newton's quotient
for $f(x) = exp(x)$.  The function should take two scalar arguments, $x$ and $h$.
Use a default value of $h=1.e-6$.
Test your function at the point $x=1$ using the default value of $h$, and compare
to the true value of the derivative $f'(1) = e^1$.


```{r}
newton_equation=function(x,h=1e-6){
    temp=(exp(x+h)-exp(x))/h
    return(temp)
}
newton_equation(1)
```
(5 points)

5.  A very useful feature in R is the ability to pass a function name as an argument.
Here is an example, where 2 is added to the value of a function, for three different
functions $exp(x)$, $log(x)$, and $sin(x)$, at selected points $x$.

```{r}
test=function(x,f){
 output=f(x)+2
 return(output)
  }

test(0,exp)
test(1,log)
test(0,sin)
test(pi/2,sin)
```

Modify your function from problem 4 so that you pass in the name of the function for which
you want to approximate the derivative.  Use the same default value for $h$, and approximate
the derivative of sin(x) at $x=\pi/4$, of $log(x)$ at $x=2$, and of $exp(x)$ at $x=1$.

```{r}
#h is the default value, 
#and should be put in the last. Otherwise it doesn't know which one is default value.
newtonf2=function(x,f,h=1e-6){
  temp=(f(x+h)-f(x))/h
  return(temp)
}
newtonf2(pi/4,sin)
newtonf2(2,log)
newtonf2(1,exp)
```
(10 points)


6. 
Write a function which takes one argument $x$ of length 2, and returns the ordered
values of $x$.  That is, if $x_2<x_1$, your function should return $c(x_2,x_1)$, otherwise it
should return $x$.  (WRITE YOUR OWN FUNCTION. DO NOT USE THE BUILT IN FUNCTION ORDER)

Use your function to process a dataset with 2 columns as follows.  Iterate over the rows of the
data set, and if the element in the 2nd column of row *i* is less than the element in the first
column of row *i*, switch the order of the two entries in the row by making a suitable call to
the function you just wrote.

Test using the following data.

```{r}
set.seed(1128719)
data=matrix(rnorm(20),byrow=T,ncol=2)
data
```


```{r}
sortf=function(x){
  if(x[2]<x[1]){
  x=c(x[2],x[1])
}
return(x)
}

sortf2=function(x){
n=nrow(x)
for (i in 1:n){
x[i,]=sortf(x[i,])
}
return(x)
}
sortf2(data)
```
(10 points)